---
title: "Broadband connection and election in Brazil: what is role of the internet?"
author: "Thiago Mendes Rosa"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    fig_height: 4
    fig_width: 6
csl: mee.csl
bibliography: references.bib
header-includes:
- \setlength\parindent{24pt}
- \usepackage{indentfirst}
- \usepackage{datetime}
- \usepackage{setspace}
- \onehalfspace
- \usepackage{pdfpages}
- \usepackage{floatrow}
- \usepackage{amsmath}
- \usepackage{morefloats}
- \usepackage{pbox}
- \usepackage{graphicx}
- \usepackage{tikz}
- \usepackage{booktabs}
- \usepackage{tabularx}
- \floatplacement{figure}{H}
- \floatsetup[figure]{capposition=top}
- \floatsetup[table]{capposition=top}
- \usepackage[bf]{caption}
- \captionsetup{justification=raggedright,singlelinecheck=false}
- \usepackage{placeins}
- \usepackage{rotating}
- \usepackage[hyphenbreaks]{breakurl}
- \usepackage{pdflscape}
mainfont: Arial
fontsize: 12pt
---

```{r setup,warning=FALSE,include=FALSE,message=FALSE,echo=FALSE}
# Load data base
load("F:/OneDrive/Doutorado/Tese/Bases/paper1.rda")

# Load packages
library(tidyverse)
library(lubridate)
library(xtable)
library(brazilmaps)
library(rdmulti)
library(rdrobust)
library(data.table)
library(rdmulti)
library(rdrobust)
library(tidyverse)
library(lubridate)
library(sf)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, dpi = 600, 
                      dev = 'png',fig.pos = 'hp')

options(xtable.comment = FALSE)

source("modelos.R",
       encoding = "UTF-8")

# Evitar confito de pacote com o kable
options(kableExtra.latex.load_packages = FALSE)
library(kableExtra)


```


# Abstract

We investigate the relationship between broadband internet and election outcomes in Brazil for 2008, 2010 and 2012. Using a robust identification strategy, a RDD aplied to the roll out of Backhaul program, we explore jumps in internet velocity according to population size. Results indicate no relationship between broadband and political outcomes -- turnout, blank and null percentage votes and left parties vote share. Our findings diverge from some results reported before, suggesting that this relationship may differ in different contexts.

# Introduction

The way how people get informed about politics has changed dramatically over the years. If in XIX century press was the main source of information, in the beginning of XX's radio took its place, surpassed by television in the middle of the same century. Today, a new type of media seems to be taking the lead: the internet.

Although the world wide web is a 25 years-old technology, broadband connection is an even recent event. Internet velocity capable of streaming videos became popular just in the XXI century. Social networks, like You Tube, Facebook and Twitter are relatively infant phenomenons, becoming popular globaly only in the late of 2000's. Mobile broadband connections, thanks to 3G technology and massification of smartphones, helped internet reach a greater number of users. New social medias, like WhatsApp and Telegram are now everyday tools, with popularity increasing in an exponential fashion, being important even for business. A new wave, with 5G technology and the "internet of things" is coming to continue the revolution begun in the past century, with connections speed and quality increasing every day.

Thus, information dissemination gained range and speed, reaching more people, almost instantaneously, nearly in any part of the world. Geographic barriers were broken and the amount of information are vast. Before these new possibilities, a question arises: how this new  scenario affects social interaction? Furthermore, how do people are doing politics in this new environment? In particular, if people have tools to be more informed, do they increase their participation in elections? Could vote preferences  change with introduction of this new technology? Or, on contrary, have this new possibilities of entertainment deviate people from political discussion? Is it possible that internet did not change at all politics?

These questions are not easy to be answered for several reasons. Availability of internet is not random and characteristics like income, schooling and geographic conditions may determine if an internet service provider will be accessible for individuals. Also, institutional and political backgrounds may possible influence internet-political relationship.

This is not a novel issue. Relationship between internet and politics has been focus of study in severel fields, and this paper aims to contribute with this literature, studying internet impacts on politics in Brazil. Studying a single country gives us best tools to control possible confounders, and using a robust identification strategy, we take advantage of a specific rule for broadband roll out that creates an instrument to deal with internet endogeneity, where number of inhabitants determines the internet velocity of municipalities (the backhaul program).

Results suggests that broadband internet are not related to political outcome in Brazil. It seems that internet did not influnce turnout, blank or null percentage votes and left parties vote share, in 2008, 2010 and 2012 elections, which covers national and local elections. The office considered (president, mayor or deputies) did not make difference in results. These finds are different from other results reported in the literature, meaning that institutional background may play an important role in political studies. Positive and negative relationship are reported for Germany, Italy and United Kingdon [@campante_politics_2017;  @falck_e-lections:_2014; @gavazza_internet_2015], all of them with distinct political institutional background. 

This paper is organized as follows: the first section presents the theoretical framework linking internet to political outcomes, while the next one reports the previous findings regarding its aplication. The third section reviews the Brazilian political background, followed by the section with empirical strategy, data bases and descriptives. The fifth section presents our results, with a final discussion in the sixth and last section.


<!-- But no other study was able to establish a causal inference considering the evolution of internet over time, both fixed line and mobile broadband, thus bringing the newest social medias to discussion. We try to cover this gap studying internet impacts in Brazil. Studying a single country gives us best tools to control possible confounders. Using a novel identification strategy, we take advantage of a specific rule for broadband rollout that creates an instrument to deal with internet endogeneity, where number of inhabitants determines the velocity of municipalities (the backhaul program). This first strategy concern fixed line broadband, while a second strategy uses 3G and 4G rollout to explore geographic and time variation to bring mobile broadband into discussion. -->


# Theoretical framework

There are some theories that try to explain why people vote [@aldrich_rational_1993; @downs_economic_1957; @ferejohn_paradox_1974; @riker_theory_1968; @uhlaner_rational_1989]. One approach is to treat as a microeconomic problem in the following way. In elections, individuals' problem is to choose the best candidate(s) according to their preferences. But, there is an asymmetry of information: there are many candidates (not considering uncontested elections), and voters are not fully informed about their habilities. Acquire information about them is costly, since they have to spend resources to consume information (e.g. from television, radio, newspaper, internet or another people), that may include money and time. Show up to cast the vote in the ballot also requires resources (transportation and time). More accurate decision requires more information, which demands more resources, i.e. is more costly. So, it can be viewed as a maximization problem from the microeconomics point of view, which can be solved by equalizing marginal costs and benefits. Benefits can be viewed as the policies the most preferred candidate will conduct, a civil duty or being party of the democratic process.

This problem changes over time with entrance of new technologies. For example, when radio, television and internet were not available, there were fewer options to people get informed about candidates. Also, there were available less leisure alternatives. With emergence of radio, then television and, finally, internet, these costs and substitution effects may have changed. A first natural question that someone could have is: did these new technologies affect the decision of voters? For newspaper, @gerber_does_2009, @gentzkow_effect_2011 and @drago_meet_2014 report effects on elections participation. According to @stromberg_radios_2004 and @horacio_media_2014, radio affects people perception about politics, while @dellavigna_fox_2007, @enikolopov_media_2011, @durante_partisan_2012, @gentzkow_television_2006 and @oberholzer-gee_media_2009 shows the impact of television (through news) on elections results. 

How about the internet? Relationship between internet and politics has been investigated since the end of 1990's [@bimber_internet_1998]. The effect on information acquisition may be ambiguous depending on the hypothesis used: if internet makes available new possibilities of entertainment, people may substitute the time spent learning about politics with these new type of leisure; on the other hand, if internet bring to people new sources of politics information and channels of discussion, people may be pushed toward politics. Finally, the cost and the time needed to find candidates information or to find new possibilities of entertainment may have changed relative prices. Once someone has access to internet, it is possible to consume a variety of information with, in general, no additional cost. The same is valid to leisure. A last possibility is that the only thing substituted is the technology used to consume information and leisure, making no difference at all in resources allocation^[If there is no, or little, consumption of politics information with an older  technology, it might be the case that, even with a new technology, there is no preference for this type of information, resulting in no, or limited, shifting in its demand.].

These changes may also take time to happen. Many types of media on internet depends on broadband connection (like video streaming), only available to the mass public in the beginning of the XXI century. Moreover, all content we have today were not available with the launch of the internet. The same was true for television, where the diversity of programs and shows existing today took time to be developed and aired. Emergency of new technologies and its spread also affects relative prices both for information and leisure over time with this development.

While newspaper, radio and TV content production are more restricted and with barrier entries, internet have opened doors to virtually anyone produce information and media, interact with people and organize groups of common interest, everything at a low cost. Thus, it is likely to exist a shift both in the demand and supply of information and entertainment with internet arrival. It can potentially alter the manner of how politics are made, since with internet politicians can reach more people quickly and at low costs.

One situation this new scenario brings is the social media consumption of fake news and its possible impact on elections. In the problem treated here, misleading information may have a market that deviate people from optimal choice [see @allcott_social_2017 for a theoretical framework]. Media capture by politicians put an additional flavor to this discussion [@besley_handcuffs_2006], where internet could break other types of media control or enhance an existing control. 

<!-- Considering that internet is a requirement for fast and wide fake news diffusion, and that its propagation depends on social media usage, assess the relationship between internet and politics seems to be the natural first step. Fake news mat be viewed as an important mechanism that can affect internet effects on politics. -->

With this framework in mind, we analyse previous researches in the field in order to collect results and identification strategies, pointing resemblances and contrasts between them. Common outcomes between internet and politics are voting turnout, election results, public polices and politician's accountability.


# Literature 

Sources from where people consume information and leisure are not exogenous. For example, if television or internet is expensive, only people with enough income can have access. If this kind of people have particular preferences regarding candidates, there is a bias if relationship between internet and politic outcomes is treated as unconditional. The same is valid for another characteristics, like race, schooling, age or housing location.

Due to this endogeneity of internet supply and demand, geographical characteristics (e.g landscape or rainfall) or previous telecommunication infrastructure are common strategies used to instrumentalize internet in order to link it to political outcomes. @campante_politics_2017 study the impact broadband diffusion on political participation for municipalities of Italy between 1996 and 2013. @miner_unintended_2015 take similar path for Malaysia, @czernich_broadband_2012 and @falck_e-lections:_2014 for Germany, @gavazza_internet_2015 for UK, @jaber_broadband_2013 for USA and @menezes_internet_2015 for Brazil. With slightly different approach, @lelkes_hostile_2017 explore variation in state laws related to internet infrastructure to study influence of this technology on polarization in USA, while @poy_internet_2016 use similar strategy to analyse broadband effects on turnout and vote share in rural and sparse areas in Italy.

For Italy, @campante_politics_2017 report a negative effect on turnout in elections following high speed internet implantation (2008), changing its direction for later elections (2013). An interesting result reported in Italian case is that internet affected ideological groups distinctly, according to vote share results, paving the way for organization of new political groups, formed in online platforms. @poy_internet_2016 echoes these results, linking high speed internet (ADSL2+) to increases in turnout in 2008 and 2013 Italian elections, as well transitory increases in vote share of some parties (center-left and right-fringe).

In Malaysian case, @miner_unintended_2015 reports important effects of internet in 2008 election results (vote share of opposition paties), but not in turnout and limited effects in turnover. It is interesting to note that the political background for the Malaysian case is different from the Italian on, although the identification strategy is similar.

A negative effect of internet on turnout is reported by @falck_e-lections:_2014 for Germany. The mechanism is related to an increase in leisure consumption that crowds out television entertainment. The impact reported is a heterogeneous one: west Germany was affect, while in east Germany no effect was observed. Effects on vote shares were not observed. On the other hand, @czernich_broadband_2012 found positive effects on participation in German 2002-2005 election.

@gavazza_internet_2015 report for UK negative effects of internet on turnout in 2006-2010 elections, with stronger results for less-educated and younger voters. Furthermore, incumbents seems to take advantage, diminishing election competitiveness. Taking a step further, the UK study suggests effects on public policies, lowering public expenses and taxes in areas with higher internet access (with similar heterogeneity effects reported for turnout).

In Brazilian case, @menezes_internet_2015 shows that internet is associated with increases in vote share of small candidates in 2010 elections, but no relation with turnout nor with no candidates votes (blank votes). This is an important result once the winner of last Brazilian presidential election (2018) won with a very limited advertisement time on radio and television in the first round.

For USA, results presented by @lelkes_hostile_2017 seems to bring light to mechanisms underlying the effects of internet one politics outcomes. States with less restrictive laws (and more likely to have broadband coverage) induces people to be exposed to partisan information and be more extreme in partisan preferences. This mechanism is compatible with results presented in @jaber_broadband_2013, who reports a positive impact on turnout, donations to political campaigns and democrats vote share in 2008 presidential elections. In an early study, with weaker identification strategy, @tolbert_unraveling_2003 suggests that, in 1996 and 2000 presidential election, individuals with internet and online elections news are more likely to vote. 

It is important to note that countries have distinct political regimes, which could potentially affect results reported. @minard_internet_2015 bring this to discussion analysing how maturity of democracy regimes in Asia responds to internet availability. Immature regimes seems to be more affect by internet than solid democracies according to 2006 cross-country analysis. Hence, this cross country variation suggests that there are institutional factors playing action on internet-politics relationship, which puts caution to external validity of results. 

To sum up, it is clear that we have different results for different countries (even inside the same country),  with possible changing effects over the time. Also, the majority of studies are concentrated in 2000 decade elections, take advantage of the begging of the internet. Few studies report results for elections held in 2010 decade, when smartphone revolution and social media gained strength. Even more, there are no studies about the offects of mobile broadband and smartphones on elections.

In this paper we will address just fixed line broadband roll out, studying the Brazilian case, one of the largest democracies in the world. Unfortunatly, mobile broadband techonology roll out (3G and 4G) will not be investigated, since information at municipality level is not available. As pointed before, peculiarities of each country seems to be determinant for results, which demands closer analysis of the political system in order to compare our results with those presented before.


# Brazil's background

Brazil is a federative republic, with three layers of government: central (or federal), states and municipalities. It is a young presidential democracy, with bicameral legislative system (Chamber of Deputies and Senate), holding election every four years. President is elected by direct vote since 1989 in national elections, being elected together with national congress, state governors and state assemblies (1994 onward). Local elections, for municipal mayors and local legislatives are also held every four years, since 1996^[Brazilian dictatorship ended in 1985, with general election in 1986, except for president (elected indirectly in the previous year). Before 1985, all other elections (except for president) had direct vote, but under military rules. In 1988, a new constitution was promulgated and in 1989 the president was elected by direct vote again, after 29 years. In 1990, there were elections for state governors, state assemblies and national congress. In 1992, municipal mayors and local assembly members were elected. By 1994 onward, national elections (president, state governors, state assembly and national congress) happens every four year, while local elections (municipal mayor and municipal assembly) happens every four years, since 1996. Thus, Brazil have elections every two years since 1994.].

In Brazil, voting is mandatory to literate citizens aged 18 to 69. For people aged 16 to 17 and over 70 voting is optional. Voters absent in election must justify or pay a small fine. If they fail to justify three consecutive pools, some rights are lost (issue or renew passports and national identification, ineligible for public education, public service and some government social programs). This set raises the question if this rule changes incentives to acquire information about politicians. 

An important aspect of Brazilian suffrage regards campaign advertisement. There are national, free of charges and mandatory programs during campaign time, booth aired daily in radio and television, broadcasting the same content in all regions of the country. There is a fixed amount of time for electoral advertisement in these channels, 2/3 distributed according to current party presence in legislatures and 1/3 among candidates, and only this time is allowed to be used in these channels. Ads on newspaper are also restricted, even though being a less important media compared to TV and radio. Internet is exception, where candidates can use, almost freely to reach voters, since 2009, except for anonymously or paid advertisement (which includes social medias like Twitter, Facebook and You Tube). 

According to @downs_economic_1957, low probability to be pivotal in elections explain the "rational ignorance" of voters and low preference to turn out. On the other hand, mandatory vote could change this incentives, making people more inclined to vote [@lijphart_unequal_1997]. @de_leon_test_2014 finds that, for Brazilian case, mandatory voting seems not change people incentivesb to be more informed in voting decision. It seems the case that providing more information about candidates [@banerjee_informed_2011], hence lowering the costs for collect information, is more effective than compulsory voting system.

Following @fujiwara_habit_2016, we try to considering possible persistent habits on voting pattern, incorporating raining information in election days in each municipality.


# Empirical strategy and data bases

## Communication usage

As a glimpse of brazilian communication consumption, Figure \ref{fig:0} presents internet and cell phone usage from 2008 to 2017.

```{r internet_usage,fig.cap="Internet and cell phone usage in Brazil, % of 16+ year-old population, 2008-2009 and 2011:2017 \\label{fig:0}"}
data.table::fread("F:/OneDrive/Doutorado/Tese/Bases/temp.csv",
                  dec=",") %>%  
  dplyr::filter(resposta=="Sim")%>%
  ggplot(aes(x=Ano,y=pct,color=var, linetype=var), size=1)+
  geom_line()+
  geom_point()+
  scale_y_continuous(labels=scales::percent)+
  scale_x_continuous(breaks = 2008:2017) +
  labs(x="Year",
       y="%",
       caption = "Source: National Household Survey (PNAD e PNADc)\nObs1.: Usage of internet in the last 3 months (survey date month as reference)\nObs2.: Cell phone ownership for personal use\nObs.3: No value for 2010 (Census year)")+
  scale_color_discrete(name="Usage of:")+
  scale_linetype_discrete(name="Usage of:")+
  theme(panel.background = element_blank(),
        axis.title = element_text(face="bold",size=12),
        axis.text = element_text(size = 10),
        legend.text = element_text(size=10),
        legend.key=element_blank(),
        plot.caption = element_text(hjust = 0,size = 8))
```

In 2008, 30% of Brazilians (16 years-old or above, i.e. population in vote age) declared to have used internet at least once in the last three months (September as reference), while almost 60% declared cell phone ownership for personal usage. In order to increase these figures, the government carried out a national plan in the beggining of 2008. In 2011, these figures rose to 44% and 73%, respectively, indicating an increasing communication market in Brazil. Even in 2017, there is room remaining for internet and cell phone expansion in Brazil.

Hence, this expressive change in communication consumption may have changed how Brazilians face politics, possibily increasing opportunities for information acquisition and social interaction about this matter. Or, on the other hand, widening leisure alternatives.

## Backhaul Program (National Broadband Plan)

In Appril 2008, the presidential Decree 6,424 changed the former National Plan of Goals for Public Switched Telephone (PST) Network Universalization, adding broandband infrastructure as mandatory (in exhange of the PST obligation). The infrastruture reefered in the Decree was the Backhaul, a requirement for internet implementation in the country. Backhauls are necessary in order to connect them to the Telephone Companies' Backbones. The plan put as target that, at least, 40% of municipalities sould have the necessary infrastructure by the end of 2008, 80% by the end of 2009 and 100% by the end of 2010. Also, minimal internet velocities were set, increasing with population size (Table \ref{tab:program_backhauk}).

```{r program_backhauk, results='asis'}
tibble(`Population Size`=c("Up to 20,000",
                               "From 20,001 to 40,000",
                               "From 40,001 to 60,000",
                               "Above 60,001"),
           `N# municipalities`=c(3077,268,63,31),
           `%`=c(89.5,7.8,1.8,0.9),
           `Velocity (Mbps)` =c(8,16,32,64)) %>%
  janitor::adorn_totals("row") %>%
  dplyr::mutate(`Velocity (Mbps)`=case_when(
    `Population Size`=="Total"~NA_real_,
    TRUE~`Velocity (Mbps)`)) %>%
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,1,0),
               caption = "Backhaul Plan",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Source: Anatel, 2010.",
                           notation="number")
```

According to the Natiaonal Agency of Telecommunication (Anatel) [@anatel_relatorio_2010], the majority of municipalities to be covered by Backhaul program were up to 20,0000 inhabitants, which is more than half of total municipalities of Brazil^[Today, Brazil has 5,570 municipalities. By the time when the program was created, six municipalities did not exist.]. The minimal required velocity (8 Mbps) guarantee improvement in navegation quality, allowing, for example, streaming (music and videos). 

Out of 5,570 municipalities, by 2015, only 85 remained uncovered (Table \ref{tab:desc_back}) and 2,125 (`r round(2125/5570*100)`%) already had broadband infrastructure before the program, mainly larger cities. We note that the program focused on small cities, with average population under 15,000.   

```{r desc_back, results='asis'}
backhaul %>%
  dplyr::filter(ano==2013) %>%
  dplyr::mutate(Situao=case_when(Situao=="Atendido"~"Covered",
                                 Situao=="Não atendido"~"Uncovered",
                                 TRUE~"Covered before")) %>%
  dplyr::group_by(Situation=Situao) %>%
  dplyr::summarise(`# Munic`=n(),
                   `Avg Velocity`=mean(backhaul,na.rm=T),
                   `Avg Pop.`=mean(popIBGE_anoantes,na.rm=T)) %>%
  janitor::adorn_totals("row") %>%
  dplyr::mutate(`Avg Pop.`=case_when(
    Situation=="Total"~mean(populacao$pop_2008,na.rm=T),
    TRUE~`Avg Pop.`))  %>%
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,rep(0,ncol(.))),
               caption = "Backhaul deployment by year",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Municipalities by backhaul status",
                           notation="number")
```

According to program schedule, 100% of Brazlis' municipalities should has backhaul infrastructure in 2010. However, by this year `r round((2125+1388+495)/5570*100)`% of the goal was achieved. Table \ref{tab:backhaul_implementation} presents the roll out of the program by year.

```{r backhaul_implementation, results='asis'}
backhaul %>%
  dplyr::filter(ano==2013,Situao=="Atendido") %>%
  dplyr::group_by(`Backhaul year`=backhaul_ano) %>%
  dplyr::summarise(`# Munic`=n(),
                   `Avg. Velocity`=mean(backhaul,na.rm=T),
                   `Avg Pop.`=mean(popIBGE_anoantes,na.rm=T)) %>%
  janitor::adorn_totals("row") %>%
  dplyr::mutate(`Avg. Velocity`=case_when(
    `Backhaul year`=="Total"~mean(backhaul[backhaul$ano==2013&backhaul$Situao=="Atendido",]$backhaul,na.rm=T),
    TRUE~`Avg. Velocity`),
    `Avg Pop.`=case_when(
    `Backhaul year`=="Total"~mean(backhaul[backhaul$ano==2013&backhaul$Situao=="Atendido",]$popIBGE_anoantes,na.rm=T),
    TRUE~`Avg Pop.`)) %>%
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,rep(0,ncol(.))),
               caption = "Backhaul deployment by year",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Source: Anatel.", 
      "Obs.: No velocity information for 2014 and 2015."),
    notation="number")
```

The main point of our identification strategy rests on the velocity descontinuity, which is futher analysed in Figures \ref{fig:1} to \ref{fig:1.5}, by region, since Brazil is a continental country with important regional inequalities. North and Northeast regions are poorer, while South and Southeast are richer^[For example, the state of São Paulo was responsible for almost 1/3 of Brazilian GDP in 2017. Per capita household income of the richest state (Federal District) was `r round(2280/593,2)` times greater than the poorest (Alagoas), according to 2014 National Household Survey (IBGE/PNAD). Brazilian Gini index for the same year was 0.517.], making a commoon practice disaggregated analysis in Brazil. 

```{r mapa, fig.cap="Internet velocity in backhaul program by municipality population, North region \\label{fig:1}",fig.height=8,fig.width=14}
# back_2008 <- backhaul %>%
#   # Retirar da análise municípios que já possuiam infraestrutura anteriormente
#   dplyr::filter(Situao %in% c("Atendido","Não atendido")) %>%
#   # Agrupar por município
#   dplyr::group_by(fk_cod_municipio) %>% 
#   # Recuperar a velocidade do backhaul para todas as linhas,
#   dplyr::mutate(backhaul=max(backhaul,na.rm=T)) %>% 
#   # Desagrupar
#   dplyr::ungroup() %>% 
#   # Ficar somente com a informação de um período.
#   # Queremos saber somente a velocidade e o ano de implementação do backhaul
#   dplyr::filter(ano==2009) %>% 
#   # Colocal como colunas o ano de implementação do backhaul com a velocidade
#   tidyr::spread(backhaul_ano, backhaul,sep="_") %>%
#   # Colocar como zero quem não teve backhaul implementado em 2008
#   # e, caso contrário, a velocidade implantada
#   dplyr::mutate(Velocity=case_when(is.na(backhaul_ano_2008)==T~0,
#                                      TRUE~as.numeric(backhaul_ano_2008)),
#                 COD_IBGE=as.character(fk_cod_municipio)) %>%
#   # Ficar somente com a informação de velocidade e o código do município
#   dplyr::select(COD_IBGE,Velocity) %>%
#   # Recuperar as informações de população
#   dplyr::left_join(populacao %>%
#                      dplyr::select(COD_IBGE,pop_2007)) %>%
#   # Renomear a variável de população
#   dplyr::rename(pop_anterior=pop_2007) %>%
#   # Retirar os municípios sem população (criados após 2007)
#   dplyr::filter(is.na(pop_anterior)==F) %>%
#   # Criar as faixas de população, os cutoffs, um marcador para internet e o ano
#   dplyr::mutate(faixa_pop=cut(pop_anterior,
#                               c(-Inf,20000,40000,60000,Inf),
#                               labels = c("Up to 20k",
#                                          "from 20k to 40k",
#                                          "from 40k to 60k",
#                                          "Above 60k"),
#                               ordered_result = T),
#                 cut_off=case_when(pop_anterior<=30000~20000,
#                                   pop_anterior %in% c(30001:50000)~40000,
#                                   pop_anterior>=50001~60000),
#                 cut_off2=case_when(pop_anterior %in% c(15000:25000)~20000,
#                                    pop_anterior %in% c(35000:45000)~40000,
#                                    pop_anterior %in% c(55000:65000)~60000),
#                 internet=case_when(Velocity>0~1,
#                                    TRUE~0),
#                 year=2008)

grafico_regiao <- function(regiao) {

  titulo <- switch(regiao, "1" = "North", "2" = "Northeast",
                   "3" = "Southeast", "4" = "South", "5" = "Center-West")

get_brmap("City") %>% 
  dplyr::mutate(City=as.character(City)) %>%
  dplyr::left_join(back_2008,
                   by=c("City" = "COD_IBGE")) %>% 
  dplyr::filter(is.na(Velocity)==F,
                Region==regiao) %>%
  ggplot() +
  geom_sf(aes(fill = Velocity),
          colour = "black", size = 0.05, alpha=0.5) +
  #muda escala de cores
  # scale_fill_viridis_d(option = 3,
  #                      direction = -1) +
  scale_fill_distiller(palette="Blues",
                       direction = 1)+
  
  geom_sf(data = filter(get_brmap("State"),Region==regiao),
          fill = "transparent",
          colour = "black", size = 0.8) +
  
  # tira sistema cartesiano
  theme(panel.grid = element_line(colour = "transparent"),
        panel.background = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = "bottom") +
  facet_wrap(~faixa_pop)+
  ggtitle(titulo)
}

# mapas <- map(1:5, grafico_regiao)
# 
# library(patchwork)
# 
# reduce(mapas, magrittr::add)

grafico_regiao(1)

```


```{r mapa2, fig.cap="Internet velocity in backhaul program by municipality population, Northeast region \\label{fig:1.2}",fig.height=8,fig.width=14}

grafico_regiao(2)

```

```{r mapa3, fig.cap="Internet velocity in backhaul program by municipality population, Southeast region \\label{fig:1.3}",fig.height=8,fig.width=14}

grafico_regiao(3)

```

```{r mapa4, fig.cap="Internet velocity in backhaul program by municipality population, South region \\label{fig:1.4}",fig.height=8,fig.width=14}

grafico_regiao(4)

```

```{r mapa5, fig.cap="Internet velocity in backhaul program by municipality population, Center-West region \\label{fig:1.5}",fig.height=8,fig.width=14}

grafico_regiao(5)
```


```{r descontinuity, fig.cap="Descontinuity in Backhaul program velocity by population cut-offs: 20,000; 40,000; 60,000 \\label{fig:2}", results='asis'}

backtemp <- backhaul %>%
  dplyr::filter(ano==2013, popIBGE_anoantes<=70000) %>%
  dplyr::mutate(cut_off=case_when(popIBGE_anoantes <=30000~20000,
                                  popIBGE_anoantes %in% c(30001:50000)~40000,
                                  popIBGE_anoantes>=50001~60000)) %>%
  na.omit

descont<-
rdmulti::rdmcplot(backtemp$backhaul,
                  backtemp$popIBGE_anoantes,
                  backtemp$cut_off,
                  pvec = c(4,4,4),
                  #hvec = c(5000,5000,5000)
                  )
```

Figure \ref{fig:2} shows a clear jump in velocity cut-offs for the entire period. The jump around cut-offs are clear, where municipalities just below the population size estabilished in the program face lower internet velocities. The first cut-off must be preferable due to sample size, as well as pooled weighted estimations, which will also be presented.

## Descriptive statistics

Despite these clear descontinuities, a set of covariates were collected, in order to control for any further confounder that might remain. Lack of information at municipal level is one of the weakness in Brazilian researchs at this territory level. Census occurs only every ten years, remaining only few administrative data in the between years, some of them with low quality (mainly for small cities). Even tough, considering that this is the only source of the main socioeconomic variables, we use information from the last two censuses (2000 and 2010), organized by Brazilian Institute of Statistics and Geography (IBGE). Also from IBGE, we collect total population estimates and GDP. Considering that direct cash transfers are important in Brazil, we collect data from the two major programs: Bolsa Família (PBF) and Benefício de Prestação Continuada for elderies (BPC), booth organized by Ministry of Citizenshipt^[PBF is one of the biggest conditional cash transfer program in the world. The target are families under the extreme poverty and poverty lines (in 2019, families earning up to R\$ 89 by person, or U\$ `r round(89/4.17)`, by month are considered extremely poor, while families above that amount and up to R\$ 178, or U\$ `r round(168/4.17)`, are considered poor), focused one children. As counter part, school attendance and vaccination are required. PBF reachs almost 14 million families in Brazil in 2019. On the other hand, BPC is a program for elderly and handicapped. The poor population in this profile (people with or over 65 years and all handicapped) are eligible for a minimum wage paycheck.]. In additon, we collect the mass of wages (formal labour market) from RAIS data base, organized by Ministry of Economy^[In Brazil, every formal company have to fill the Annual Relation of Social Information (RAIS), with the profile of all workers they had in the calendar year, including wages.].We also collected information from National Insitute of Metereology, to control for rain and temperature in election day, following @fujiwara_habit_2016. Municipalities were joined by the nearest distance between the center of the city and the closest meteorological station. 

In addition, we collected data from Ministry of Health regarding homicidy and suicide, but, due to poor quality of data (missing values), we had to discard them. Public finance data were collected too, but discarted due to missing problem and for being highly correlated with other covariates (like GDP and population). In fact, some of covariates were discarted due to high correlation (for example eletricity, car and computer ownership). Finally, we collected fiscal data, from Ministry of Economy, regarding public expenses (current expenses and investments), but again due to missing data and high correlation with other covariates (like GDP and mass of wages) these variables were dropped too. 

Outcomes are election results, organized by Superior Election Court (TSE). We will analyze 2008, 2010 and 2012 elections, covering two municipal elections and one national election. The main outcomes we will analyze are: turnout, percentage of blank or null votes^[In Brazilian election, people may put a blank vote, which are not computed for any candidate and is not considered for official results, as well null votes. The difference consists in the way the registration of these votes are made: the blank vote is available as a button in the electronic ballot, while the null vote occurs when someone enters an invalid candidate number into the ballot and confirms the vote.] and vote shares, for left wing, center or right wing parties (following @power_elite_2012 and the party index from Brazilian Legislative Survey^[Version 7, available in https://dataverse.harvard.edu/dataverse/bls;jsessionid=992eedb7e954a17ef718c7078cf5?widget=dataverse%40harvard&q=&types=dataverses%3Afiles%3Adatasets&sort=dateSort&order=desc&page=3]. Left wing parties are those up to quantile 0.25 of the party index, center parties are those between 0.25 and 0.75 and right wing parties are those above.). 

Descriptive statistics are separated by year (2008, 2010 and 2012), considering 2000 Census data for 2008 statistics and 2010 Census for 2010 and 2012. All the other variables refers to the respective year.

```{r desc.2008, results='asis'}
# source_lines <- function(file, lines){
#     source(textConnection(readLines(file,
#                                     encoding = "UTF-8")[lines]))
# }
# 
# source_lines("F:/OneDrive/Doutorado/Tese/tese_1/modelos.R",
#              11:480)
# 
# source_lines("F:/OneDrive/Doutorado/Tese/tese_1/modelos.R",
#              808:845)

descriptives_2008 %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,rep(2,4)),
               caption = "Descriptive Statistics, 2018",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Source: IBGE, Inmet, ME and MC",
    notation="number")
```

```{r desc.2010, results='asis'}
# source_lines("F:/OneDrive/Doutorado/Tese/tese_1/modelos.R",
#              847:884)

descriptives_2010 %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,rep(2,4)),
               caption = "Descriptive Statistics, 2010",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Source: IBGE, Inmet, ME and MC",
    notation="number")
```


```{r desc.2012, results='asis'}
# source_lines("F:/OneDrive/Doutorado/Tese/tese_1/modelos.R",
#              886:923)

descriptives_2012 %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,rep(2,4)),
               caption = "Descriptive Statistics, 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Source: IBGE, Inmet, ME and MC",
    notation="number")

```

All three tables shows a high heterogeneity in Brazil. Regional inequalities, as pointed before, are particularly important and should be considered in standard error estimates. It is very likely that municipalities in the northeast and in the south regions show different behavior that may possibly affect error distributions, that should be addressed by, for example, clustered standard errors.

In order to clarify identification validity, Table \ref{tab:t.test} presents a simple t-test for 20,000 population with a 5,000 people (adhoc) cutoff. Since this is the cutoff with larger sample size, results should be more robust.

```{r t.test, results='asis'}
# source_lines("F:/OneDrive/Doutorado/Tese/tese_1/modelos.R",
#              608:695)

t_base %>% 
  dplyr::rename_all(list(~c("Variable","t 2008","p value 2008",
                            "t 2010","p value 2010",
                            "t 2012","p value 2012"))) %>% 
    dplyr::mutate(Variable=c("Turnout",
                             "Population",
                             "Median Income",
                             "Pop. over 60 years",
                             "Rural",
                             "Black",
                             "Radio",
                             "Television",
                             "College",
                             "Married",
                             "Working Pop.",
                             "Rain (elect. day)",
                             "Avg. Temperature",
                             "PBF",
                             "BPC",
                             "GDP",
                             "Formal Wages",
                             "Velocity")) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,rep(3,6)),
               caption = "Covariates t-test for 20,000, 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote("Source: IBGE, Inmet, ME and MC",
    notation="number")

```

Results for 2008 year in Table \ref{tab:t.test} (which, indeed, refers to 2000 census for socioeconomic variables) show that there were no significant differences for most of the characteristics between municipalities just above and just below the cut-off, except for the population (and velocity), as expected. Some results, however, do not hold in 2010 and 2012, years with values collected from 2010 census and, hence, closer to the years of analysis. Some covariates, like formal wages and BPC seems to be different across municipalities, while other covariates changes its significance in 2010 sample and 2012. Overall, \ref{tab:5} results suggests that our identification strategy should work, if controled for some covariates, and even more if applied to a narrow cutoff (which will be the case).

# Results

We begin our analysis of results looking the effects of broadband in participation. Considering that there are two rounds for two types of offices, mayor and president, and some municipalities might not have a second round, we focus only in the first one, using the larger sample size as possible. All the three cut-offs are considered (20,000, 40,000 and 60,000), with optimal bandwidth chosen following @imbens_optimal_2012. Regressions are performed in R software, with rdd package^[Drew Dimmery (2016). rdd: Regression Discontinuity Estimation. R package version 0.57. https://CRAN.R-project.org/package=rdd] and rdmulti package^[Matias D. Cattaneo, Rocio Titiunik and Gonzalo Vazquez-Bare (2018). rdmulti: Analysis of RD Designs with Multiple Cutoffs or Scores. R package version 0.2. https://CRAN.R-project.org/package=rdmulti], for pooled regression with multiple cutoffs, following @cattaneo_analysis_2018.

Results in Table \ref{tab:reg_part} suggest no relationship between broadband internet and participation in elections. For all regressions, considering the three cutoffs, and pooled results, only two showed significant relationship. Since participation in elections is mandatory in Brazil, and turnout is relatively high (around 80%), there is no much room to improve this situation. These results are in line with @menezes_internet_2015. Another way to look to this relationship is the difference in turnout between elections, presented in Table \ref{tab:reg.part.dif}. 

```{r reg.part, results='asis'}
r_part %>% 
  dplyr::distinct() %>% 
  dplyr::arrange(Year, Cutoff,Type) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression results for turnout. Election years: 2008, 2010 and 2012. Brazil",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions",
      "Participation for first round",
      "Bw=bandwidth",
      "LATE estimates."),
    notation="number")
```

Results suggests that broadband did not changed turnout rate between elections in Brazil. It is an interesting result since a new possibility of leisure, at least apparently, did not reduce people participation in elections. However, in another backgrounds, where participation in elections are not mandatory, results may be different (like in Italy case showed by @campante_politics_2017).

```{r reg.part.dif, results='asis'}
r_dif_part %>% 
  dplyr::distinct() %>% 
  dplyr::arrange(Year, Cutoff,Type) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression results for difference in turnout. Election years: 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions.",
      "Participation for first round of elections.",
      "Bw=bandwidth.",
      "LATE estimates."),
    notation="number")
```

The next outcome regards to the percentage of blank or null votes (Table \ref{tab:r.pct.bn}). Again, there is little support fo the influence of broadband internet in blank or null votes (a *proxy* for "absence of engagement with political process", since these votes can be seen as a "whatever vote"). So, results so far suggests that broadband did not encourage or discourage people to turnout neither people to place more directed votes in elections.

```{r r.pct.bn, results='asis'}
r_pct_bn %>% 
  dplyr::distinct() %>% 
  dplyr::arrange(Year, Cutoff,Type) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression results for blank of null votes. Election years: 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions.",
      "For mayor elections (2008 and 2012) and presidential election (2010).",
      "Bw=bandwidth.",
      "LATE estimates."),
    notation="number")
```

No huge changes are observed when we look to the differences in these percetages between elections (Table \ref{tab:r.pct.bn.dif}).

```{r r.pct.bn.dif, results='asis'}
r_bn_dif %>% 
  dplyr::distinct() %>% 
  dplyr::arrange(Year, Cutoff,Type) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression results for blank of null votes. Election years: 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions.",
      "For mayor elections (2008 and 2012) and presidential election (2010).",
      "Bw=bandwidth.",
      "LATE estimates."),
    notation="number")
```

In last presidential elections (2018), polarization was dramatic in Brazil. Left versus Right debate was at the center of the presidential debate, with the last four times presidential winner party (the left wing Workers Party -- PT) being the main target. In fact, 2014 elections was one of the closest seen in Brazil, when Mrs. Rouseff defeated Mr. Neves (from central right Brazilian Social Democracy Party -- PSDB) with only 51.64% of the valid votes in the second round. Internet may had a important role in this scenario, since, back in 2010, Mr. Lula da Silva, the first president of Workers Party, had 80% of presidency approval, the highest value ever recorded^[A news about these figures are available in: http://g1.globo.com/politica/noticia/2010/12/popularidade-de-lula-bate-recorde-e-chega-87-diz-ibope.html].

Hence, a closer look on the relationship between broadband and vote share of left parties since 2008 might shed light into this turnaround in Brazil. As pointed before, vote shares were classified as left, center or right based on @power_elite_2012 party index. Table \ref{tab:parties} presents this organization.

```{r parties, results='asis'}
dplyr::tibble(Left=c("PSTU","PSOL","PC do B",
                     "PT","PSB","PCO",
                     "","","","","","",""),
              Center=c("PDT","PV","PCB","PPS","PSDB","PMDB",
                       "PTB","PSD","PL","PRONA","PR","PSC","PRB"),
              Right=c("PRN","PFL","DEM","PDS","PPR","PDC",
                      "PPB","PP","PMN","PSL","","","")) %>% 
  knitr::kable("latex",
               booktabs = T,
               caption = "Party classification based on party index",) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Division of parties based on quantiles of party index (0.25, 0.75, 1)",
      "Parties out of party index were allocated based on party description available on their internet page."),
    notation="number")
```

The party index has some aggregation of parties as "others", so another classification criterion was necessary. Parties web pages were consulted to analyse their history and beliefs in order to designate parties to the groups. This methodology may arise questions if some parties labeled as right are actually centrists. To avoid this issue, we focus on left parties vote shares, since their classification are more direct and mostly based on the party index.

Results suggests, once again, that there is no clear relationship between broadband internet and the vote received by left wing parties in elections for mayors and president (Table \ref{tab:r.pct.vote}). 

```{r r.pct.vote, results='asis'}
r_pct_vote %>% 
  dplyr::distinct() %>% 
  dplyr::arrange(Year, Cutoff,Type) %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression for left wing parties vote share. Election years: 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position")) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions.",
      "For mayor elections (2008 and 2012) and presidential election (2010).",
      "Left wing parties: PSTU, PSOL, PC do B, PT, PSB and PCO.",
      "Bw=bandwidth.",
      "LATE estimates."),
    notation="number")
```

A limitation of RDD models is the bandwidth choice, that could influence results. It is possible that a narrower or wider bandwidth give different results, since fewer or more observation will be part of regressions (trade off between randomness and precision). Considering this possibility, Table \ref{tab:r.sig} presents only significant results using also half or double bandwidths.

```{r r.sig,results='asis'}
r_sig %>% 
  knitr::kable("latex",
               booktabs = T,
               digits = c(0,0,0,0,0,3,3,3,3),
               caption = "RDD regression with alternatibe bandwidths. Election years: 2008, 2010 and 2012",
               format.args = list(decimal.mark = '.', big.mark = ",")) %>%
  kableExtra::kable_styling(latex_options = c("hold_position"),
                            font_size = 10) %>%
  kableExtra::add_footnote(
    c("Standard Errors are clustered by regions.",
      "For mayor elections (2008 and 2012) and presidential election (2010).",
      "Left wing parties: PSTU, PSOL, PC do B, PT, PSB and PCO.",
      "Bw=bandwidth.",
      "LATE estimates."),
    notation="number") %>%
  kableExtra::landscape()

```

With bandwidth doubled, a significant and negative effect of broadband are observed for left wing vote share in 2008 (for the first and last cutoffs). However, results are only significant with covariates, a possible consequence due to the loss of randomness due to the increased distance from the cut-off. Turnout seems to be positively related to broadband around 40,000 cut-off, while negatively related around 60,000, this last one more robust to covariates and changes in the cut-off. This is an interesting result, suggesting that cities with different sizes may respond differently to the internet. 

In 2010, negative relationship between broadband and left wing vote shares seems to persist around 20,000 cutoff, although there is an ambiguity in a specification with covariates, changing its sign in LATE regression from positive to negative when covariates are omitted. Considering that around the 60,000 cut-off results are positive with covariate, these figures should be looked with even more caution. Turnout shows a positive relationship only with half bandwidth, for 20,000 and 40,000 cutoffs, which don't seems to be a very robust results because they are sensitive to covariates. Black and null percentage votes are only significant for 20,000 cutoff, being highly sensitive to covariates. Booth LATE and doubled bandwidth regression show a change in sign when covariates are considered, putting in check the results.

Finally, left wing vote share is positively related in the first cut-off and negatively related in the last one, but only without covariates. Turnout shows a change of sign in the 20,000 cutoff when the bandwidth is narrowed, making any conclusion problematic. Finally, blank and null votes shows distinct behavior according to the cut-off, being negatively related to broad band in the first and the last one, while positively related in the middle.

Putting all these results together, it is hard to conclude that the backhaul program, and, hence, broadband availability, made difference in 2008, 2010 and 2012 elections in terms of turnout, left wing vote share and percentage of blank or null votes, at least when presidential and mayor offices, and only the first round, area considered^[These are the natural scenarios to be first investigated because they are the more important offices in each election.]. It may possible that other offices present a different result. However, results (not reported) are similar, with significance only for a negative relationship between broadband and left wing vote share in municipal chamber office election (vereador), only for the first cut-off. One more time, significant results are sparse, which means that it is hard to point a consistent pattern linking broadband internet availability to election outcomes.

# Discussion and conclusions

Relationship between broadband and elections outcome does not seems be relevant in Brazil, at least when landline broadband are considered, neither for local or national elections. Despite our robust identification strategy, we did not find strong relationship between broadband availability, measured by the jumps of internet velocity in the backhaul program roll out, and election outcomes (turnout, percentage of blank and null votes and left parties vote share).

These results are different from those reported in the literature, which could indicate that the background may be important in this kind of analysis. First of all, vote is mandatory in Brazil, which is not necessarily true in other countries. Second, the political system in Brazil is presidentialist, in a federative republic, which means that people may behavior differently than in a parlamentarism system. Third, national congress deputies are elected by proportional vote, while senators are elected by direct vote, situation that may differ across countries. A fourth source of variation in political background regards to the difference between unitary and federal systems, that sets different rules to be played.

Aside the political background, there is a qualification of internet usage that is not addressed in our analysis. First of all, social networks have grown in Brazil after 2010. Whatsapp, one of the most popular social network in Brazil today, was created only in 2009, the very same year internet campaing was regulated. Is it possible that, today, mobile broadband and social medias usage in smartphones are more important for communication and mobilization than older social medias and connection made at home, through desktop computers. Unfortunately, there is not available in Brazil the roll out of 3G and 4G technology implementation at municipality level, only at Direct Dialling codes (DDDs)^[The DDD codes are numbers that divides Brazil in `r length(unique(smp2009_2018$DDD))` areas.], which makes impossible to determine when the technology begun to operate in every city^[We contact the Regulation Agency of Telecomunication -- Anatel -- requesting mobile internet implementation at municipality level. Unfortunatly, there is no such data base available.].

Nonetheless, our paper contributes to bring into discussion that internet and political outcomes should be viewed in a wider perspective, meaning that some relationship may be circunstantial to idiosyncrasies of the countries. Also, futher investigation, like the role of the new social medias and mobile broadband, are necessary to shed light in this discussion, even because internet and social medias are still evolving.

# References